On the probability of finding a nonphysical solution through shadowing

### 2020-10-29 main.tex
Change final figure

Through a rigorous counter-example, we show that shadowing can lead
to nonphysical solutions. This has a troubling implication for numerical simulations of chaotic
governing equations.  Through the examples in this paper, it is clear that a numerical solution, satisfying
a governing equation perturbed due to numerics, is not 
expected to be shadowed by a
physical solution of the real governing physics.  What is, if any, the
relation of a numerical solution, such as DNS of turbulent flows, to
the true physics? For an ${\cal O}(10^{-18})$ 
perturbation (due to numerical error), 
we may get completely different statistics from a true, physical solution. How then can we trust numerical solutions when they are not guaranteed to share the long-term statistical behavior of the governing equation? 
We conclude with an example that illustrates a counterintuitive feature of some chaotic systems: a small perturbation to the 
governing equation can significantly change the statistics/long-term 
behavior of its physical solutions. We construct yet another 
perturbed tent map -- the \emph{plucked} 
tent map -- in which an oscillatory perturbation is introduced (see Supplementary Material section \ref{sec:pluckedTentMap} for the map equation). In Figure \ref{fig:pluckedTentMap}, we show that the magnitude of the oscillatory perturbation is controlled by parameters $s$ and $n$, and its frequency is controlled by the parameter $n$; at $s=0$, we recover the original tent map at all $n$. The stationary probability distributions -- computed empirically over a trajectory of length 10 billion -- at $s=0.1$, are shown on the right column of Figure \ref{fig:pluckedTentMap}. On the top row, $n=0$, and there is already a marked asymmetry developed in the probability distribution compared to the uniform distribution seen at $s=0, n=0$. The figure shows that by increasing $n,$ despite the fact that the magnitude of the perturbation becomes smaller with $n$, we see a dramatic change in the appearance of the stationary probability distribution: we observe an apparent fractal distribution reminiscent of that of the quasi-physical solutions in 
section \ref{sec:tentPerts}. By construction of the map (Supplementary Material section \ref{sec:pluckedTentMap}), the nonuniformities of the probability distribution at $n=0$ are transferred to smaller and smaller scales, 
as $n$ is increased. It is worth emphasizing that these remarkable changes in the physical distribution (shown on the bottom-right of Figure \ref{fig:pluckedTentMap}) are effected with a tiny perturbation -- the perturbed map at $n=6$, and the original tent map appear indiscernible on the bottom-left of Figure \ref{fig:pluckedTentMap}.

In view of the plucked tent map, we can question further the
validity of numerical solutions. Can numerical play 
the role of the plucking perturbation to the 
governing equation? That is, can numerical solutions represent 
slight perturbations to the governing equations that nevertheless 
exhibit drastically different statistics from that of the ensemble? Uniformly hyperbolic 
dynamical systems exhibit \emph{linear response} 
(\cite{ruelle}\cite{baladi}) by which small parameter perturbations 
lead to small changes in statistics, which can be expanded as Taylor series around the reference parameter value. But, uniform hyperbolicity is a mathematical idealization, and although some physical systems have been observed to behave as if they were uniformly hyperbolic \cite{gallavotti}, a violation of uniform hyperbolicity is more likely 
\cite{wormell}\cite{luca}. Sauer \cite{sauer} has shown 
examples in which numerical error due to computation in double-precision floating point arithmetic causes significant change in the stationary probability distribution on the attractor. The plucked tent map adds to this list of examples by exhibiting a mechanism to produce an extreme non-smooth response (Supplementary Material section \ref{sec:pluckedTentMap}).

@@

The supplementary material for this paper including the code, data and figures can be found on Github \cite{supp}. The code can be found under the \verb+code+ subdirectory inside of which section-wise code is separated into further subdirectories. The data used to plot the figures can be found under \verb+data+. The files referred to in this section can be found in the appropriate subdirectory under \verb+code+. 

@@

closed form expression for the map is necessary for our numerical shadowing procedure. In a small region around the cusp, we approximate the map using an exponential function. The tails on both sides are fitted with a sum of a cubic polynomial and a rational function. Thus, the approximate Lorenz map has the 

@@

in the Lorenz map iterates are denoted $z_{\rm max}$ and $z_{\rm min}$, respectively. The exponent of the cusp is denoted $p$. The coefficients of the cubic polynomial modelling the left (right) tail are  denoted $p_{L,n}, n = 0,1,2,3$ ($p_{R,n}, n = 0,1,2,3$). The coefficients of the numerator and denominator of the rational function modelling the left (right) tail are denoted $a_{L,0}, a_{L,1}$ ($a_{R,0}, a_{R,1}$) and $b_{L,n}, n=0,1,2,3$ ($b_{R,n}, n=0,1,2,3$), respectively. The values of these coefficients are obtained by 
regression on the map at standard parameters are shown in the table below. As shown in Figure \ref{fig:lorenzMapComp}, the fit obtained matches the Lorenz map closely.

@@

    \caption{The fitting parameters of the Lorenz map at $\sigma = 10, \beta = 8/3$ and $\rho = 28.$}

@@

\subsection{Computing shadowing solutions}
\label{sec:computingShadowing}
		In general, to numerically compute the shadowing solutions, one could use existing methods such as 
		the least squares shadowing method \cite{qiqi-lss}. However, since these maps are all one-dimensional chaotic systems that have in common non-invertibility with two inverse \emph{branches}, we devise a simpler approach. The map $\varphi_s$ in this section refers to any of the perturbations of the tent map (section \ref{sec:tentPerts}) or the Lorenz map. Suppose we are given a perturbed solution $x_n, n =0,1,\cdots,N$ that we must compute a shadow of. 
The shadowing solution, $y_n, n=0,1,2,\cdots,N$ 
must a) satisfy the governing equation: $y_{n+1} = \varphi_s(y_n)$
and b) lie close to the given perturbed solution at all times up to 
$N$, i.e., $|x_n - y_n| < \epsilon$, for some $\epsilon > 0$, for all $n \leq N$. Suppose we set $y_0 = x_0$, the difference $y_1 - x_1$ will amplify on further iterations under $\varphi_s$ for any perturbed solution. On the other hand, backward iteration of the map is contracting, and thus we set $y_N = x_N$ and proceed backward in time to 
construct a solution of $\varphi_s$, noting that any difference that emerges at a given time $n$ will be made smaller starting at $n-1$. Each point has two pre-images under $\varphi_s$, and thus there are two possible choices for $y_{N-1}$, each lying in one of two fixed sub-intervals separated by the cusp of $\varphi_s$. Due to the contraction of errors backward in time, as long as we choose a pre-image in the same subinterval as the perturbed solution, we are guaranteed to approximate a shadowing solution. Marching backward by choosing at each step, the pre-image $y_n$ in the same subinterval as $x_n$, $y_n$ approximates the shadowing solution better as $n$ decreases. Thus, the procedure to find a shadowing solution simply reduces to solving for a backward trajectory (specifically one among the possible $2^N$), starting at a given final condition $x_N$.

@@

\subsection{The \emph{plucked} tent map}
\label{sec:pluckedTentMap}
We provide a recursive definition of the plucked tent map, which 
is illustrated at different values of $n$ and $s$ in Figure \ref{fig:pluckedTentMap}. First we define a function $f_s(x)$, which 
creates a bend in the tent map that increases with $s$, around 
$((1-s)/2,1)$:
\begin{align}
		f_s(x) = 		{\rm min}\Big(
				\dfrac{2x}{1-s}, 2 - \dfrac{2(1-x)}{1+s}\Big),
				\;\;\; x < 1 \\
\end{align}
Then, we introduce oscillations using repetitions of
the above map within the unit interval,
\begin{align}
		o_s(x) = \begin{cases}
				f_s(2x)/2, & \; x < 0.5 \\
				2 - f_s(2-2x)/2, & \; x \ge 0.5.
		\end{cases}
\end{align}
We can modulate
the frequency or repetitions -- proportional to $n$ -- 
of the oscillations through
\begin{align}
		\label{eqn:frequency}
		\lambda_{s,n}(x) = \dfrac{o_s(2^n x - \floor*{2^n x})}{2^n} 
		+ 2 \frac{\floor*{2^n x}}{2^n},
\end{align}
where $\floor*{x}$ is the greatest integer less than or equal to $x$.
Finally, the plucked tent map is defined as the above function 
in the unit interval, and as its reflection about $x=1$,
in the interval $[1,2)$.
\begin{align}
		\label{eqn:plucked}
		\varphi_{s,n}(x) = {\rm min}\left(
		\lambda_{s,n}(x), \lambda_{s,n}(2-x)\right), \; 0< x < 2. 
\end{align}
Recall that our motivation is to construction a slight perturbation of the tent map whose stationary probability distribution is not just 
non-uniform (e.g. like the other tent map perturbations of section \ref{sec:tentPerts}) but in which the nonuniformity can be controlled, and made skewed as desired. Thus, we choose to construct a base perturbation of the tent map, ${\rm min}(o_s(x), o_s(2-x))$ -- 
which is also obtained by setting $n=0$ in $\varphi_{s,n}$ --  
in which an asymmetry is produced in the probability distribution (top-right of Figure \ref{fig:pluckedTentMap}).
Upon repeating, with appropriate scaling, the oscillation $o_s$, 
the frequency of which is controlled by $n$, we obtain $\varphi_{s,n}$, 
as indicated by Eq. \ref{eqn:frequency}-\ref{eqn:plucked}. Thus, the asymmetric probability distribution, through this process of scaled repetition, acquires an apparent fractal-like structure seen on increasing $n$, as shown on the bottom-right of Figure \ref{fig:pluckedTentMap}.

One way to see the asymmetry in the probability distribution 
about $x=1$, at $n=0$, is to construct a Markov chain with nodes 00, 01, 10, and 11, that indicate the subintervals $(0,0.5)$, $[0.5,1)$, 
$[1,1.5)$ and $[1.5,2)$, respectively. From Eq. \ref{eqn:plucked}, we can see that the application of $\varphi_{s,0}$ yields the following transition matrix:
\[
\begin{bmatrix}
		(1-s)/2\quad & (1+s)/2 \quad & 0 \quad & 0 \\
		0 \quad & 0 \quad & (1+s)/2 \quad & (1-s)/2 \\
		0 \quad & 0  \quad & (1+s)/2 \quad & (1-s)/2 \\
		(1-s)/2 \quad & (1+s)/2 \quad & 0 \quad & 0
\end{bmatrix}
.\] 
The stationary probability distribution of this Markov chain, which 
is the left eigenvector of the transition matrix corresponding to eigenvalue 1, is $[(1-s)^2/2, \; (1-s^2)/2, \; (1+s)^2/2,\; (1-s^2)/2]^T$.
One can verify this is consistent with the top-right of Figure \ref{fig:pluckedTentMap} for $\varphi_{0.1,0}$. Clearly, the time spent by an infinitely long physical solution in the second quarter interval, $[0.5,1]$ is 1.2 times the time spent in the first quarter, $[0,0.5]$; the time spent in the third quarter $[1,1.5]$ is 1.5 times that spent in the first quarter. This nonuniformity is more pronounced with increasing $s$. For instance, at $s=0.5$, the left subinterval, [0,1),is three times less likely to be visited by a long trajectory compared to $[1,2)$.

But, as mentioned in the main text, even a small perturbation -- very small $s$ -- is sufficient to trigger a remarkable variation in the statistics. This is because this particular construction transfers 
the nonuniformities in the probability distribution at larger scales to smaller scales, with increasing $n.$ The nonuniformity at the largest 
scale of half intervals is retained at higher values of $n.$ That is,
the probability of visiting the interval $[0,1]$ is the same: $(1-s)$, at all $n$. However, with increasing $n,$ the nonuniformity emerges within smaller subintervals, due to the construction of the map that relies on repeating the behavior at larger scales at smaller scales (Eq. \ref{eqn:frequency}). This repetitive construction is responsible for the apparent fractal structure of the probability distribution (Figure \ref{fig:pluckedTentMap}, bottom-right). The scripts that generate the plucked tent map and its 
stationary probability distributions can be found under \verb+tent_sens_stat+. 
\end{document}

>>>>

Through a rigorous counter-example, we show that shadowing can lead
to nonphysical solutions. This has a troubling implication for numerical simulations of chaotic
governing equations.  Through the examples in this paper, it is clear that a numerical solution, satisfying
a governing equation perturbed due to numerics, is not 
expected to be shadowed by a
physical solution of the real governing physics.  What is, if any, the
relation of a numerical solution, such as DNS of turbulent flows, to
the true physics? For an ${\cal O}(10^{-18})$ 
perturbation (due to numerical error), 
we may get completely different statistics from a true, physical solution. How then can we trust numerical solutions when they are not guaranteed to share the long-term statistical behavior of the governing equation? 
We conclude with an example that illustrates a counterintuitive feature of some chaotic systems: a small perturbation to the 
governing equation can significantly change the statistics/long-term 
behavior of its physical solutions. We construct yet another 
perturbed tent map -- the \emph{plucked} 
tent map -- in which an oscillatory perturbation is introduced (see Supplementary Material section \ref{sec:pluckedTentMap} for the map equation). In Figure \ref{fig:pluckedTentMap}, we show that the magnitude of the oscillatory perturbation is controlled by parameters $s$ and $n$, and its frequency is controlled by the parameter $n$; at $s=0$, we recover the original tent map at all $n$. The physical probability distributions -- computed empirically over a trajectory of length 10 billion -- at $s=0.1$, are shown on the right column of Figure \ref{fig:pluckedTentMap}. On the top row, $n=0$, and there is already a marked asymmetry developed in the physical distribution compared to the uniform distribution, which is the physical distribution seen at $s=0, n=0$. The figure shows that by increasing $n,$ despite the fact that the magnitude of the perturbation becomes smaller with $n$, we see a dramatic change in the appearance of the stationary probability distribution: we observe an apparent fractal distribution reminiscent of that of the quasi-physical solutions in 
section \ref{sec:tentPerts}. By construction of the map (Supplementary Material section \ref{sec:pluckedTentMap}), the nonuniformities of the probability distribution at $n=0$ are transferred to smaller and smaller scales, 
as $n$ is increased. It is worth emphasizing that these remarkable changes in the physical distribution (shown on the bottom-right of Figure \ref{fig:pluckedTentMap}) are effected with a tiny perturbation -- the perturbed map at $n=6$ and the original tent map appear indiscernible on the bottom-left of Figure \ref{fig:pluckedTentMap}.

In view of the plucked tent map, we can question further the
validity of numerical solutions. Can numerical solutions play 
the role of the plucking perturbation to the 
governing equation? That is, can numerical solutions represent 
slight perturbations to the governing equations for which the physical 
distribution is drastically different from the physical distribution 
of the original governing equation? Sauer \cite{sauer} has shown 
examples in which numerical error due to computation in double-precision floating point arithmetic causes significant change in the stationary probability distribution on the attractor. The plucked tent map adds to this list of examples by demonstrating a mechanism to produce an extreme non-smooth response (Supplementary Material section \ref{sec:pluckedTentMap}). Uniformly hyperbolic 
dynamical systems exhibit \emph{linear response} 
(\cite{ruelle}\cite{baladi}) by which small parameter perturbations 
lead to small changes in statistics, which can be expanded as Taylor series around the reference parameter value. But, uniform hyperbolicity is a mathematical idealization, and although some physical systems have been observed to behave as if they were uniformly hyperbolic \cite{gallavotti}, a violation of uniform hyperbolicity is more likely 
\cite{wormell}\cite{luca}.

@@

The supplementary material for this paper including the code, data and figures can be found on Github \cite{supp}. The code can be found under the \verb+code+ subdirectory inside which section-wise code is separated into further subdirectories. The data used to plot the figures can be found under \verb+data+. The files referred to in this section can be found in the appropriate subdirectory under \verb+code+. 

@@

closed form expression for the map is necessary for our numerical shadowing procedure (section \ref{sec:computingShadowing}). In a small region around the cusp, we approximate the map using an exponential function. The tails on both sides are fitted with a sum of a cubic polynomial and a rational function. Thus, the approximate Lorenz map has the 

@@

in the Lorenz map iterates are denoted $z_{\rm max}$ and $z_{\rm min}$, respectively. The exponent of the cusp is denoted $p$. The coefficients of the cubic polynomial modelling the left (right) tail are  denoted $p_{L,n}, n = 0,1,2,3$ ($p_{R,n}, n = 0,1,2,3$). The coefficients of the numerator and denominator of the rational function modelling the left (right) tail are denoted $a_{L,0}, a_{L,1}$ ($a_{R,0}, a_{R,1}$) and $b_{L,n}, n=0,1,2,3$ ($b_{R,n}, n=0,1,2,3$), respectively. The values of these coefficients, which are obtained by
regression, are shown for the map at standard parameters, in the table below. As shown in Figure \ref{fig:lorenzMapComp}, the fit obtained matches the Lorenz map closely.

@@

    \caption{The fitting parameters of the Lorenz map at $\sigma = 10, \beta = 8/3$, and $\rho = 28.$}

@@

\subsection{Computing shadowing solutions}
\label{sec:computingShadowing}
		In general, to numerically compute the shadowing solutions, one could use existing methods such as 
		the least squares shadowing method \cite{qiqi-lss}. However, since the  maps we consider are all one-dimensional chaotic systems that have in common non-invertibility with two inverse \emph{branches}, we devise a simpler approach. The map $\varphi_s$ in this section refers to any of the perturbations of the tent map (section \ref{sec:tentPerts}) or the Lorenz map. Suppose we are given a perturbed solution $x_n, n =0,1,\cdots,N$ that we must compute a shadow of. 
The shadowing solution, $y_n, n=0,1,2,\cdots,N$ 
must a) satisfy the governing equation: $y_{n+1} = \varphi_s(y_n)$
and b) lie close to the given perturbed solution at all times up to 
$N$, i.e., $|x_n - y_n| < \epsilon$, for some $\epsilon > 0$, for all $n \leq N$. Suppose we set $y_0 = x_0$, the difference $y_1 - x_1$ will amplify on further iterations under $\varphi_s$ for any perturbed solution. On the other hand, backward iteration of the map is contracting, and thus we set $y_N = x_N$. Then, we proceed backward in time to 
construct a solution of $\varphi_s$, noting that any difference that emerges at a given time $n$ will be made smaller starting at $n-1$. Each point has two pre-images under $\varphi_s$, and thus there are two possible choices for $y_{N-1}$, each lying in one of two fixed sub-intervals separated by the cusp of $\varphi_s$. Due to the contraction of errors backward in time, as long as we choose a pre-image in the same subinterval as the perturbed solution, we are guaranteed to approximate a shadowing solution. Marching backward by choosing at each step, the pre-image $y_n$ in the same subinterval as $x_n$, $y_n$ approximates the shadowing solution better as $n$ decreases. Thus, the procedure to find a shadowing solution simply reduces to solving for a backward trajectory (specifically one among the possible $2^N$), starting at a given final condition,  $x_N$.

@@

\subsection{The \emph{plucked} tent map}
\label{sec:pluckedTentMap}
We provide a recursive definition of the plucked tent map, which 
is illustrated at different values of $n$ and $s$ in Figure \ref{fig:pluckedTentMap}. First we define a function $f_s(x)$, which 
creates a bend in the tent map that increases with $s$, around 
$((1-s)/2,1)$:
\begin{align}
		f_s(x) = 		{\rm min}\Big(
				\dfrac{2x}{1-s}, 2 - \dfrac{2(1-x)}{1+s}\Big),
				\;\;\; x < 1. 
\end{align}
Then, we introduce oscillations using repetitions of
the above map within the unit interval,
\begin{align}
		o_s(x) = \begin{cases}
				f_s(2x)/2, & \; x < 0.5 \\
				2 - f_s(2-2x)/2, & \; x \ge 0.5.
		\end{cases}
\end{align}
We can modulate
the frequency of repetitions -- proportional to $n$ -- 
of the oscillations through
\begin{align}
		\label{eqn:frequency}
		\lambda_{s,n}(x) = \dfrac{o_s(2^n x - \floor*{2^n x})}{2^n} 
		+ 2 \frac{\floor*{2^n x}}{2^n},
\end{align}
where $\floor*{x}$ is the greatest integer less than or equal to $x$.
Finally, the plucked tent map is defined as the above function 
in the unit interval, and as its reflection about $x=1$,
in the interval $[1,2)$.
\begin{align}
		\label{eqn:plucked}
		\varphi_{s,n}(x) = {\rm min}\left(
		\lambda_{s,n}(x), \lambda_{s,n}(2-x)\right), \; 0< x < 2. 
\end{align}
Recall that our motivation is to construction a slight perturbation of the tent map whose stationary probability distribution is not just 
non-uniform (e.g. like the other tent map perturbations of section \ref{sec:tentPerts}), but in which the nonuniformity can be controlled, and made skewed as desired. Thus, we choose to construct a base perturbation of the tent map, ${\rm min}(o_s(x), o_s(2-x))$ -- 
which is also obtained by setting $n=0$ in $\varphi_{s,n}$ --  
in which an asymmetry is produced in the probability distribution (top-right of Figure \ref{fig:pluckedTentMap}).
Upon repeating, with appropriate scaling, the oscillation $o_s$, 
the frequency of which is controlled by $n$, we obtain $\varphi_{s,n}$, 
as indicated by Eq. \ref{eqn:frequency}-\ref{eqn:plucked}. Thus, the asymmetric probability distribution, through this process of scaled repetition, acquires an apparent fractal-like structure seen on increasing $n$, as shown on the bottom-right of Figure \ref{fig:pluckedTentMap}.

One way to see the asymmetry in the probability distribution 
about $x=1$, at $n=0$, is to construct a Markov chain with nodes 00, 01, 10, and 11, that indicate the subintervals $(0,0.5)$, $[0.5,1)$, 
$[1,1.5)$ and $[1.5,2)$, respectively. From Eq. \ref{eqn:plucked}, we can see that the application of $\varphi_{s,0}$ yields the following transition matrix:
\[
\begin{bmatrix}
		(1-s)/2\quad & (1+s)/2 \quad & 0 \quad & 0 \\
		0 \quad & 0 \quad & (1+s)/2 \quad & (1-s)/2 \\
		0 \quad & 0  \quad & (1+s)/2 \quad & (1-s)/2 \\
		(1-s)/2 \quad & (1+s)/2 \quad & 0 \quad & 0
\end{bmatrix}
.\] 
The stationary probability distribution of this Markov chain, which 
is the left eigenvector of the transition matrix corresponding to eigenvalue 1, is $[(1-s)^2/2, \; (1-s^2)/2, \; (1+s)^2/2,\; (1-s^2)/2]^T$.
One can verify this is consistent with the top-right of Figure \ref{fig:pluckedTentMap} for $\varphi_{0.1,0}$. Clearly, the time spent by an infinitely long physical solution in the second quarter interval, $[0.5,1]$ is about 1.2 times the time spent in the first quarter, $[0,0.5]$; the time spent in the third quarter $[1,1.5]$ is about 1.5 times that spent in the first quarter. This nonuniformity is more pronounced with increasing $s$. For instance, at $s=0.5$, the left subinterval, [0,1),is three times less likely to be visited by a long trajectory compared to $[1,2)$.

But, as mentioned in the main text, even a small perturbation -- very small $s$ -- is sufficient to trigger a remarkable variation in the statistics. This is because this particular construction transfers 
the nonuniformities in the probability distribution at larger scales to smaller scales, with increasing $n.$ The nonuniformity at the largest 
scale of half intervals is retained at higher values of $n.$ That is,
the probability of visiting the interval $[0,1]$ is the same: $(1-s)$, at all $n$. However, with increasing $n,$ the nonuniformity emerges within smaller subintervals, due to the construction of the map that relies on repeating the behavior at larger scales at smaller scales (Eq. \ref{eqn:frequency}). This repetitive construction is responsible for the apparent fractal structure of the probability distribution (Figure \ref{fig:pluckedTentMap}, bottom-right). The scripts that generate the plucked tent map and its 
physical probability distributions can be found under \verb+tent_sens_stat+. 
\end{document}
